Remote sensing riverine gas exchange and carbon efflux using SWOT observations

### 2021-02-17 manuscript_v1.Rmd
manuscript edits

---
title: "Remote sensing riverine gas exchange from SWOT and implications for estimating carbon efflux from rivers"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
csl: american-geophysical-union.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Key Points
- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite and no in situ information
- BIKER can nearly replicate estimates of $CO_2$ fluxes made via hydraulic geometry and an in situ streamgauge
- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
- BIKER's accuracy is limited by process-level uncertainties in gas exchange from rivers and not by its algorithmic setup

##Abstract
Go BIKER!!
x
x
x
x
x
x
x
x
x
x

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976). Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

@@

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the [@ulsethDistinctAirWater2019a] dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is heavily informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following [@tsivoglouTracerMeasurementReaeration1976], $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area *dA*). It is also defined by the upscaling parameters $a_i$ and $b_i$.

@@

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification each term in equation 5 is log-transformed such that equation 5 is a lognormal distribution, however they are written here without a transformation for succinctness. $σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).

@@

Equation five necessitates that Bayesian priors be assigned to model parameters n, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020]. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

@@

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity. To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity. Estimate uncertainty for velocity followed [@hagemannBAMBayesianAMHGManning2017a], who estimated ln-transformed Manning’s equation error to be 0.25. This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.

With the sampling model (equation 5) and prior distributions described, a joint posterior distribution conditional on the SWOT data is derived. To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other MCMC approaches (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived in [@frassonCompilationHydraulicModels2019a] and used by [@rodriguezObservingRiversVarying2020] to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

We also introduced measurement error into these simulated SWOT observations to assess the performance degradation due to errors inherent in SWOT measurements. Error in SWOT-measured river heights and water surface slopes will come from tolerances internal to the satellite, i.e. the SWOT error budget, as well as radar layover error. River width errors are assumed here to be neglible [@durandHowWillRadar2020a]. Layover error is the phenomenon when radar returns arrive at the sensor at the same time, generally in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013]. While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as defined as $S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2})$ and $H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2})$ (in meters). Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to upscale $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets). Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is calculated using equation 7, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2.

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(7)}$$

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE). Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a] and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate *V*.

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors. The names used here refer to the final implementation used in each study’s upscaling exercises. For this study, all rating-curve methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models)). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year, i.e. multiplied by the surface area of all 22 rivers.

@@

Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations (section 2.2) for the 8,000 hydraulic measurements mapped in Figure 4c. Figure 4a plots a histogram of the 8,000 uncertainty terms extracted from those distributions (Figure 4b plots three of the distributions used to build Figure 4a). Across the 8,000 tests, we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$meanSigma,2)` (~`r round(exp(results_3_1_MC$meanSigma), 2)` m/day- dashed blue line in Figure 4a). Uncertainty solely from the upscaling model is 1.30, confirming that using Manning’s equation to approximate average flow velocity is introducing very little additional uncertainty to the model (Figure S2). This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by [@hallGasExchangeStreams2020].

![Figure 4 Model uncertainty estimates: a: histogram of SDs for all 8,000 MC simulations with blue dashed line denoting the mean (1.32). b: Three example MC simulation results of 10,000 samples each. c: Map of the 8,000 sets of hydraulic measurements used for this analysis.](outputs/MonteCarlo/MCsimulations.jpg)

###3.2 Validating the BIKER algorithm
Figure 5a plots the validation results for $k_{600}$ (with no measurement error) across all 22 rivers and all timesteps. For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions. $k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER, capturing the general magnitude of the predictions and with most points falling on or near the 1:1 line. However, the 95% prediction intervals (dashed grey lines) highlight relatively large residuals for many predictions. Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions. This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that recovering upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.

@@

Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b. Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the 95% CIs for the BIKER posterior. We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated below the median $k_{600}$ values and systemically overestimated above the median $k_{600}$ However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b). This highlights the benefit of using a Bayesian predictive framework to fully propagate all parameter uncertainties through to the posterior. While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers. Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the large uncertainties inherent in our $k_{600}$ estimates.

>>>>

---
title: "Remote sensing riverine gas exchange and carbon efflux using SWOT observations"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Key Points (85 characters each)
- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
- BIKER replicates estimates of $CO_2$ fluxes when coupled with in situ concentration data
- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup

##Abstract
Go BIKER!!
400 words
x
x
x
x
x
x
x
x
x

@@

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

@@

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, using data from [@ulsethDistinctAirWater2019a] dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem.

@@

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ for the change in channel area *dA* and median channel area $A_0$). It is also defined by the upscaling parameters $a_i$ and $b_i$.

@@

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness. $σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).

@@

Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020]. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

@@

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. We specifically explore three uncertainty scenarios to assess which error source is most influential. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we ran three sets of MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The specifics of the slope-error model used are elaborated in Section 2.3, but in short we account for both the SWOT error tolerances and radar layover error [@durandHowWillRadar2020a]. This resulted in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.

With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is derived. To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

We also introduced measurement error into these simulated SWOT observations to assess the performance degradation due to errors inherent in SWOT measurements. Error in SWOT-measured river heights and water surface slopes will come from tolerances internal to the satellite, i.e. the SWOT error budget, as well as radar layover error. River width errors are assumed here to be neglible [@durandHowWillRadar2020a]. Layover error is the phenomenon when radar returns arrive at the sensor at the same time, generally in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013]. While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as by equations 7 and 8. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.

$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) (7)$$

$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) (8)$$

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets). Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is calculated using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2.

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(9)}$$

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE). Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a].

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors. The names used here refer to the specific implementation used in each study. For our study, all HG methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year from the 22 rivers, accounting for total river surface area.

@@

Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations for three error scenarios (section 2.2) for the 5,000 hydraulic measurements. Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms). Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a). Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measuurement error). This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by @hallGasExchangeStreams2020.

![Figure 4 Model uncertainty estimates for three error scenarios, as detailed. Left: histogram of error terms (SD) for all 5,000 MC simulations with blue dashed line denoting the median. Right: Three example MC simulation results of 10,000 samples each.](outputs/MonteCarlo/MC_results.jpg)

###3.2 Validating the BIKER algorithm
Figure 5a plots the validation results for $k_{600}$ (with no measurement error) across all 22 rivers and all timesteps. For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions. $k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER, capturing the general magnitude of the predictions and with most points falling on or near the 1:1 line. However, the 95% prediction intervals (dashed grey lines) highlight relatively large residuals for many predictions. Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions. This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that inferring upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.

@@

Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b. Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the 95% CIs for the BIKER posterior. We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated below the median $k_{600}$ values and systemically overestimated above the median $k_{600}$ However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b). This highlights one benefit of using Bayesian inference to fully propagate all prior uncertainties through to the posterior. While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers. Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the large uncertainties inherent in our $k_{600}$ estimates.
